---
output: github_document
bibliography: bibliography.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file and build with devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bfsep

<!-- badges: start -->
<!-- badges: end -->

The baseflow separation model `bfsep` is a wrapper library around the [code and tools](https://www.sciencebase.gov/catalog/item/5f90ef0282ce720ee2d29b7e) built by Christopher Konrad at USGS [@Konrad2020] and is used for the estimation of the baseflow component of streamflow and is executed as a set of 12 functions. The model is run using the function `bf_sep()`, which requires six arguments.  All other functions are called by `bf_sep()`.  All required functions can be loaded in the library:

```{r}
library(bfsep)
```


The model has been calibrated for `r nrow(bf_params_usgs)` selected gaging sites. Parameters are available for the USGS gage parameters dataset `bf_params_usgs` included with this package, as excerpted below.  Description of the parameters in this dataset can be viewed by calling the help `?bf_params_usgs`.

```{r}
library(magrittr)
#library(flextable)

# flextable(tail(bf_params_usgs)) %>% 
#   colformat_num(big.mark=",", digits = 2, na_str = "N/A") %>% 
#   autofit()

knitr::kable(tail(bf_params_usgs), digits=2, format.args = list(scientific = FALSE))
```

Sites are in rows with identification number in the first column, the 16 parameters values, and the error for the calibration. 

The parameters for a site must be concatenated into 3 vectors with six elements each before using them as arguments to `bf_sep()`.  Helper functions for each set exist to facilitate retrieving the gage-specific parameters.

columns 2:6 for basin_char, columns 7:11 for gw_hyd, and columns 12:17 for flow.

`bfsep` works best in conjunction with the USGS `dataRetrieval` package to fetch time-series data from specific gages.

## Installation

You can install the development version of `bfsep` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jvandens/bfsep")
```

## Example

This is a basic example which shows how to download discharge data from a single USGS site and calculate the baseflow separation

```{r example, warning=FALSE}

library(dataRetrieval)

# SAUK RIVER NEAR SAUK, WA
siteNumber <- "12189500"
parameterCd <- "00060"  #discharge in cfs

# Raw daily data:
rawDailyData <- readNWISdv(
  siteNumber,
  parameterCd,
  startDate = "2000-01-01",
  endDate = "2000-12-31")
```

Rename the columns to 'friendly' names and the `bfsep` model requires flow to be in m3/day so add a conversion column as default returned is in cfs

```{r}
# rename to friendly names and convert to m3/day
rawDailyData <- renameNWISColumns(rawDailyData) %>% 
  dplyr::mutate(Flow_m3 = Flow / 35.3147 * 3600 * 24)
```

Plot the data as a quick check
```{r}

library(ggplot2)

ggplot(rawDailyData, aes(x=Date, y=Flow_m3)) +
  geom_line()
```

bfsep has helper functions to lookup the parameters for a site which are required for the model.  The gage-specific parameters required are the basin characteristics (`basin_char()`), groundwater hydrology (`gw_hyd()`), and flow parameters (`flow()`).  Each of these functions takes the USGS gage ID character string as it's single argument.

```{r}

# model args
basin_char <- basin_char(siteNumber) 
gw_hyd <- gw_hyd(siteNumber)
flow <- flow(siteNumber) 
timestep <- 'day' 
error_basis <- 'total' 

# prepare the gage data as vectors
dt <- rawDailyData$Date #VECTOR OF DATES
qin <- rawDailyData$Flow_m3 #VECTOR OF DAILY STREAMFLOW VALUES IN M3/day
```

Run the baseflow separation model, save the results to a dataframe and output the associated error

```{r}
sep <- bf_sep(qin, dt, timestep, error_basis, basin_char, gw_hyd, flow)
```

A basic plot of the results:

```{r}

sep.plt <- sep %>% 
  dplyr::select(Date:DirectRunoff.L3) %>% 
  tidyr::pivot_longer(2:6, names_to = "var", values_to = "Flow_m3/day") 

ggplot(sep.plt, aes(x = Date, y = `Flow_m3/day`, color = var)) +
  geom_line(show.legend = FALSE) +
  facet_wrap(~var, ncol=1)

```

# References

<!--
You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>. 
-->
